%% ============================================================
%%  Golden Substrate series -- canonical preamble.
%%  Identical across all papers except for \hypersetup{pdftitle=...}.
%% ============================================================

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage{xcolor}

\definecolor{linkInternal}{HTML}{1F4E79}
\definecolor{linkExternal}{HTML}{2F5496}
\definecolor{linkCite}{HTML}{806600}
\usepackage[colorlinks=true,
            linkcolor=linkInternal,
            citecolor=linkCite,
            urlcolor=linkExternal,
            pdfborder={0 0 0},
            bookmarksnumbered=true,
            breaklinks=true]{hyperref}
\hypersetup{%
  pdfauthor={Kristin Tynski},
  pdftitle={Working Outline: Z[phi] Apollonian Thin Groups and Level-11 Hauptmoduln},
  pdfsubject={Copyright (c) 2026 Kristin Tynski. All rights reserved.}%
}

\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[protrusion=true,expansion=false]{microtype}
\usepackage{fancyhdr}
\usepackage[font=small,labelfont=bf,format=plain,labelsep=period]{caption}
\usepackage[framemethod=tikz]{mdframed}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc,decorations.pathmorphing,shapes.geometric,fit,backgrounds}

\geometry{margin=1in}
\setlength{\emergencystretch}{5em}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{principle}[theorem]{Principle}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\numberwithin{equation}{section}

\providecommand{\Cl}{\mathrm{Cl}}
\providecommand{\R}{\mathbb{R}}
\providecommand{\C}{\mathbb{C}}
\providecommand{\Z}{\mathbb{Z}}
\providecommand{\Q}{\mathbb{Q}}
\providecommand{\N}{\mathbb{N}}
\providecommand{\F}{\mathbb{F}}
\providecommand{\HH}{\mathbb{H}}
\providecommand{\OO}{\mathbb{O}}
\providecommand{\Zp}{\ensuremath{\mathbb{Z}[\varphi]}}
\providecommand{\Qp}{\ensuremath{\mathbb{Q}(\sqrt{5})}}
\providecommand{\Gphi}{\ensuremath{\Gamma_{\varphi}}}
\providecommand{\vp}{\varphi}
\providecommand{\ps}{\psi}
\providecommand{\eps}{\varepsilon}
\providecommand{\half}{\tfrac{1}{2}}
\providecommand{\fk}[1]{\mathfrak{#1}}
\providecommand{\cO}{\mathcal{O}}
\DeclareMathOperator{\disc}{disc}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Aut}{Aut}
\providecommand{\id}{\mathrm{id}}

\definecolor{accentGray}{HTML}{6B7C93}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\small\itshape\color{accentGray}\nouppercase{\leftmark}}
\fancyhead[R]{\small\color{accentGray}\thepage}
\fancyfoot[C]{}
\renewcommand{\sectionmark}[1]{\markboth{\thesection.\ #1}{}}

\widowpenalty=1000
\clubpenalty=1000
\setlength{\parskip}{0.55em plus 0.2em minus 0.05em}
\setlength{\parindent}{1.2em}

\title{Working outline: $\mathbb{Z}[\varphi]$ Apollonian thin groups and level-$11$ Hauptmoduln}
\author{Kristin Tynski}
\date{May 2026 \\[6pt] {\small\textcopyright\ 2026 Kristin Tynski. All rights reserved.}}

\begin{document}
\maketitle

\noindent\textbf{Place in the series.}
This is a working scaffold appended to \emph{Paper~IV (Moonshine
context)} of the Golden Substrate series.  It is not a finished
paper; it is the postponed long-form treatment of the
$\mathbb{Z}[\varphi]$ Apollonian thin-group programme and is
preserved here so that the Track~1 brute outcome and the planned
section layout remain reachable from the public repo.

\section*{Status}

This skeletal file anchors the postponed long-form manuscript envisaged once
automatic comparisons between truncated $\mathbb{Z}[\varphi]$ Apollonian
orbits and Conway--Norton series mature beyond brute histogram matching.

\textbf{Recorded Track~1 brute outcome (engineering oracle):}

\begin{enumerate}
\item Norm-keyed multiplicity series from both classical Galois-fixed and
birth-unit irrational seeds mismatch the McKay-Thompson series for class \texttt{11A}
coefficient-wise wherever overlaps are tested---windowed audits show zero coincidences
under the naive literal comparison harness.
\item Atkin $w_{11}$ symmetrising code paths remain unimplemented stubs.
\item Dual-embedding symmetric products remain heuristic probes only (integer pair counters
implement a bivariate norm generating function analogue in code).
\end{enumerate}

\textbf{Branches for future exposition}

\begin{enumerate}
\item \textit{literal match attained under twist} \(\Rightarrow\) document transformation law + sibling moonshine;
\item \textit{consistent twisted invariant} \(\Rightarrow\) name intermediate Hilbert object;
\item \textit{persistent obstruction} \(\Rightarrow\) publish negative empirical result with refined conjecture.
\end{enumerate}

\subsection*{Planned sections (body text TBD)}
\begin{itemize}
\item Analytic prelude: Mayer generators plus Hao--Qin--Zhou uniqueness datum.
\item Orbit taxonomy: Galois-fixed lifts vs irrational seeds, norm bounds, maximal-quadruple truncation.
\item Generating-function normalisations (Jacobian vs diagonal restrictions).
\item Speculative lemmas linking thin subgroup congruence data to Hauptmoduln quotients.
\item Appendix bridging classical Apollonian modularity obstructions (Haag et al.\ 2024) with reciprocity folklore.
\end{itemize}

\section*{Remark}
Populate a full bibliography once the branching narrative stabilises after Track~2B authoring.

\end{document}
