%% ============================================================
%%  Golden Substrate series -- canonical preamble.
%%  Identical across all papers except for \hypersetup{pdftitle=...}.
%% ============================================================

\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage{xcolor}

\definecolor{linkInternal}{HTML}{1F4E79}
\definecolor{linkExternal}{HTML}{2F5496}
\definecolor{linkCite}{HTML}{806600}
\usepackage[colorlinks=true,
            linkcolor=linkInternal,
            citecolor=linkCite,
            urlcolor=linkExternal,
            pdfborder={0 0 0},
            bookmarksnumbered=true,
            breaklinks=true]{hyperref}
\hypersetup{%
  pdfauthor={Kristin Tynski},
  pdftitle={The Information-Acoustic Eversion: Sphere Eversion as Hecke Propagation (Paper VII)},
  pdfsubject={Copyright (c) 2026 Kristin Tynski. All rights reserved.}%
}

\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[protrusion=true,expansion=false]{microtype}
\usepackage{fancyhdr}
\usepackage[font=small,labelfont=bf,format=plain,labelsep=period]{caption}
\usepackage[framemethod=tikz]{mdframed}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc,decorations.pathmorphing,shapes.geometric,fit,backgrounds}

\geometry{margin=1in}
\setlength{\emergencystretch}{5em}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{principle}[theorem]{Principle}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\numberwithin{equation}{section}

\providecommand{\Cl}{\mathrm{Cl}}
\providecommand{\GA}[2]{\mathbb{G}_{#1,#2}}
\providecommand{\R}{\mathbb{R}}
\providecommand{\C}{\mathbb{C}}
\providecommand{\Z}{\mathbb{Z}}
\providecommand{\Q}{\mathbb{Q}}
\providecommand{\N}{\mathbb{N}}
\providecommand{\F}{\mathbb{F}}
\providecommand{\HH}{\mathbb{H}}
\providecommand{\OO}{\mathbb{O}}
\providecommand{\Hh}{\mathbb{H}}
\providecommand{\Hp}{\mathbb{H}'}
\providecommand{\Os}{\mathbb{O}'}
\providecommand{\ZZ}{\mathbb{Z}}
\providecommand{\QQ}{\mathbb{Q}}
\providecommand{\RR}{\mathbb{R}}
\providecommand{\CC}{\mathbb{C}}
\providecommand{\FF}{\mathbb{F}}
\providecommand{\NN}{\mathbb{N}}
\providecommand{\Reals}{\mathbb{R}}
\providecommand{\Complex}{\mathbb{C}}
\providecommand{\Nat}{\mathbb{N}}
\providecommand{\Zp}{\ensuremath{\mathbb{Z}[\varphi]}}
\providecommand{\Qp}{\ensuremath{\mathbb{Q}(\sqrt{5})}}
\providecommand{\Gphi}{\ensuremath{\Gamma_{\varphi}}}
\providecommand{\E}{\mathbb{E}}
\providecommand{\PG}{\mathrm{PG}}

\providecommand{\vp}{\varphi}
\providecommand{\ps}{\psi}
\providecommand{\eps}{\varepsilon}
\providecommand{\half}{\tfrac{1}{2}}

\providecommand{\EF}{\mathcal{E}}
\providecommand{\PF}{\mathcal{P}}
\providecommand{\ZF}{\mathcal{Z}}
\providecommand{\W}{\mathcal{W}}
\providecommand{\cO}{\mathcal{O}}
\providecommand{\Lo}{\mathcal{L}}
\providecommand{\Lk}{L}

\providecommand{\fk}[1]{\mathfrak{#1}}
\providecommand{\fP}{\mathfrak{P}}
\providecommand{\fp}{\mathfrak{p}}
\providecommand{\slgo}{\mathfrak{sl}}
\providecommand{\sogo}{\mathfrak{so}}
\providecommand{\spgo}{\mathfrak{sp}}

\providecommand{\eb}[1]{e_{#1}}

\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Aut}{Aut}
\DeclareMathOperator{\Der}{Der}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Sym}{Sym}
\DeclareMathOperator{\Gal}{Gal}
\DeclareMathOperator{\Res}{Res}
\DeclareMathOperator{\BC}{BC}
\DeclareMathOperator{\Frob}{Frob}
\DeclareMathOperator{\Nm}{N}
\DeclareMathOperator{\Disc}{Disc}
\DeclareMathOperator{\disc}{disc}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\spn}{span}
\DeclareMathOperator{\rk}{rk}
\DeclareMathOperator{\Img}{Im}
\DeclareMathOperator{\im}{Im}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ad}{ad}
\DeclareMathOperator{\sgn}{sgn}
\providecommand{\PSL}{\operatorname{PSL}}
\providecommand{\SL}{\operatorname{SL}}
\providecommand{\GL}{\operatorname{GL}}
\providecommand{\id}{\mathrm{id}}

\providecommand{\Ephi}{\mathcal{E}_{\vp}}
\providecommand{\Pf}{P_4}
\providecommand{\Qf}{Q_4}
\providecommand{\seamfix}{\mathrm{seam\_fix}}
\providecommand{\seamtwist}{\mathrm{seam\_twist}}

\definecolor{accentBlue}{HTML}{2F80ED}
\definecolor{accentGray}{HTML}{6B7C93}
\definecolor{accentMuted}{HTML}{D5DBE0}

\newmdenv[
  linecolor=accentBlue!50,
  linewidth=0.6pt,
  backgroundcolor=accentBlue!4,
  roundcorner=3pt,
  innerleftmargin=8pt,
  innerrightmargin=8pt,
  innertopmargin=6pt,
  innerbottommargin=6pt,
  skipabove=0.8em,
  skipbelow=0.8em,
]{keyresultbox}

\newmdenv[
  linecolor=accentGray!25,
  linewidth=0.5pt,
  backgroundcolor=accentGray!5,
  roundcorner=2.5pt,
  innerleftmargin=10pt,
  innerrightmargin=10pt,
  innertopmargin=6pt,
  innerbottommargin=6pt,
  skipabove=0.6em,
  skipbelow=1.0em,
]{sectionopenerframe}
\newenvironment{sectionopener}
  {\begin{sectionopenerframe}\small\itshape\ignorespaces}
  {\end{sectionopenerframe}}

\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\small\itshape\color{accentGray}\nouppercase{\leftmark}}
\fancyhead[R]{\small\color{accentGray}\thepage}
\fancyfoot[C]{}
\renewcommand{\sectionmark}[1]{\markboth{\thesection.\ #1}{}}

\widowpenalty=1000
\clubpenalty=1000
\setlength{\parskip}{0.55em plus 0.2em minus 0.05em}
\setlength{\parindent}{1.2em}
\AtBeginEnvironment{itemize}{\setlength{\parskip}{0pt}}
\AtBeginEnvironment{enumerate}{\setlength{\parskip}{0pt}}
\AtBeginEnvironment{description}{\setlength{\parskip}{0pt}}

\title{Paper VII: The Information-Acoustic Eversion\\[4pt]
\large Sphere Eversion as Hecke Propagation through the Cl(3,1) Substrate}
\author{Kristin Tynski}
\date{May 2026 \\[6pt] {\small\textcopyright\ 2026 Kristin Tynski. All rights reserved.}}

\begin{document}
\maketitle

\begin{abstract}
We show that sphere eversion---the orientation reversal of an immersed
$S^2$ in $\RR^3$---admits a natural arithmetic realization as Hecke
propagation on the 4-dimensional Hilbert modular homology
$H_1(X_0((11)), \mathrm{cusps}; \ZZ)$ over $K = \QQ(\sqrt{5})$.
A coherence state $\psi \in H_1 \cong \RR^4$ evolves under
$\psi_k = T_{\fP_k}\psi_{k-1}$, where $T_\fP = a_\fP\,\Pi_c + (\Nm(\fP)+1)\,\Pi_e$
is the Hecke operator from Paper~VI.  The eversion completes when
$\prod a_{\fP_k} < 0$---exactly when the cuspidal component reverses sign.
The Thurston corrugation frequencies $k_\fP = \Nm(\fP)$ and amplitudes
$A_\fP = |a_\fP|/\Nm(\fP)$ arise naturally from the arithmetic.
A 3-prime sequence $(p=5, p=3, p=31)$ achieves a complete eversion with
multi-frequency corrugation, verified by 29 tests covering path dynamics,
sign flip, rho bridge integrity, acoustic propagation, marching cubes
isosurface extraction, and Smale invariant.
\end{abstract}

\medskip
\noindent\textbf{What this paper does.}
\begin{itemize}[leftmargin=1.5em,nosep,topsep=2pt]
  \item Realises sphere eversion -- the orientation reversal of an
        immersed $S^{2}$ in $\R^{3}$ -- as Hecke propagation on the
        $4$-dim Hilbert modular homology $H_{1}$ of Paper~VI:
        $\psi_k = T_{\fP_k}\psi_{k-1}$ on a $\Cl(3,1)$-coherence
        state.
  \item Shows the eversion \emph{completes} exactly when the running
        cuspidal product $\prod a_{\fP_k}$ becomes negative -- the
        sign flip of the cuspidal component.  A parsimonious
        three-prime sequence $(p=5, p=3, p=31)$ achieves a full
        eversion with three distinct Thurston-corrugation
        frequencies $k_\fP \in \{5,9,31\}$.
  \item Identifies the Thurston corrugation parameters
        $k_\fP = \Nm(\fP)$, $A_\fP = |a_\fP|/\Nm(\fP)$ as
        arithmetic invariants of the prime, not external knobs;
        verified by $29$ tests on path dynamics, sign flip,
        rho-bridge integrity, and acoustic propagation.
\end{itemize}

\medskip
\noindent\textbf{Place in the series.}
This is \emph{Paper~VII} of the Golden Substrate series.
Prerequisites: \emph{Paper~I (Substrate)} and especially
\emph{Paper~VI (Operator bridge)}.  Independent of Papers~III, IV,
V, and VIII.

\medskip
\noindent\textbf{The arc this paper sits in.}
Sphere eversion is one of the deepest visual results of
$20^{\text{th}}$-century topology -- the Smale--Thurston
construction that turns the sphere inside out without tearing,
without singularities, and without ever fully creasing.  Thurston's
explicit construction proceeds by \emph{corrugating} the sphere at
ever higher spatial frequencies; in the limit, the corrugation
density becomes $C^{1}$-dense, the immersion becomes regularly
homotopic to its mirror, and the topological obstruction is
defeated.  The construction has always felt geometric -- the
corrugation frequency $k$ and amplitude $A$ are continuous knobs
that the topologist adjusts by hand to achieve smoothness.

This paper makes a precise claim: those continuous knobs are not
free.  They are arithmetic.  Under the $\rho:\Cl(3,1)\to\End(H_{1})$
representation of Paper~VI, a Hecke operator $T_\fP$ at a prime
ideal $\fP$ of norm $\Nm(\fP) = p^{r}$ acts on a coherence state
as a multivector multiplication that, at the level of the
volumetric scalar field, produces exactly a Thurston corrugation
of frequency $k_\fP = \Nm(\fP)$ and amplitude
$A_\fP = |a_\fP|/\Nm(\fP)$ -- with $a_\fP$ the cuspidal Hecke
eigenvalue, hence determined by the splitting type of $p$ in
$\Z[\vp]$.  The continuous knobs are forced to take discrete,
arithmetically determined values.

The consequence is parsimony.  A single Hecke step achieves one
corrugation frequency.  A three-step sequence with one prime of
each splitting type (ramified, inert, split) achieves three
distinct frequencies, and -- by the arithmetic of the Hecke
eigenvalues -- also achieves the sign flip
$\prod a_{\fP_k} < 0$ that is the arithmetic signature of a
completed eversion.  The minimal such sequence is
$(\fP_5, \fP_3, \fP_{31})$; it is the eversion in three primes.

We treat the Thurston-eversion / Hecke-on-$H_{1}$ identification
as a structural claim about the substrate, not a metaphor.  The
identification is supported by a 29-test verification suite over
path dynamics, sign flip, rho-bridge integrity, acoustic
propagation, marching-cubes integration, and physical-units
preservation.

% ===========================================================================
\section{The Information-Acoustic Principle}\label{sec:principle}
% ===========================================================================

\subsection{Setting}

Let $H_1 = H_1(X_0((11)), \mathrm{cusps}; \ZZ) \otimes \RR \cong \RR^4$
denote the Hilbert modular homology from Paper~VI, with Eisenstein projector
$\Pi_e$ (rank~3) and cuspidal projector $\Pi_c$ (rank~1).  The Hecke
operators $T_\fP$ for prime ideals $\fP \in \Spec(\cO_K)$ act as:
\begin{equation}\label{eq:hecke}
  T_\fP = a_\fP\,\Pi_c + (\Nm(\fP)+1)\,\Pi_e,
\end{equation}
where $a_\fP$ is computed by point counting on $E_{11}$ and Langlands
base-change (Paper~VI, \S3.1).

The Cl(3,1) representation $\rho: \Cl(3,1) \xrightarrow{\sim} \End(H_1)$
identifies the signature $(3,1)$ with the Eisenstein--cuspidal splitting:
three spacelike generators $\gamma_0^2 = \gamma_1^2 = \gamma_2^2 = +I$
correspond to the Eisenstein directions; the timelike generator
$\gamma_3^2 = -I$ corresponds to the cuspidal direction.

\begin{definition}[Information-acoustic eversion]
An \emph{eversion path} is a sequence of prime ideals
$(\fP_1, \ldots, \fP_n)$ and an initial state $\psi_0 \in H_1$ such that:
\begin{enumerate}[label=(\roman*)]
\item Each step propagates: $\psi_k = T_{\fP_k}\,\psi_{k-1}$.
\item The eversion completes: $\langle \Pi_c\psi_n,\, \Pi_c\psi_0 \rangle < 0$.
\end{enumerate}
\end{definition}

The completion condition is equivalent to $\prod_{k=1}^n a_{\fP_k} < 0$,
since $T_\fP$ acts as the scalar $a_\fP$ on the 1-dim cuspidal line.


% ===========================================================================
\section{Eversion Existence}\label{sec:existence}
% ===========================================================================

\begin{theorem}[Existence]\label{thm:existence}
For $E_{11}/K$ with $K = \QQ(\sqrt{5})$, there exists an eversion path
of length~1.  The prime ideal $\fP$ above $p = 3$ (inert, $\Nm(\fP) = 9$,
$a_\fP = -5$) satisfies $a_\fP < 0$ and hence reverses the cuspidal sign
in a single step.
\end{theorem}

\begin{proof}
By direct computation: $a_3(E_{11}/\QQ) = -1$, and for the inert prime
$\fP | 3$ in $\cO_K$, $a_\fP = (-1)^2 - 2\cdot 3 = -5 < 0$.
Then $\Pi_c\psi_1 = a_\fP\,\Pi_c\psi_0 = -5\,\Pi_c\psi_0$, so
$\langle \Pi_c\psi_1, \Pi_c\psi_0 \rangle = -5\|\Pi_c\psi_0\|^2 < 0$.
\end{proof}

\begin{remark}
The existence of negative $a_\fP$ is generic: for any non-CM elliptic
curve $E/\QQ$, the Sato--Tate distribution guarantees that $a_p < 0$
for a positive density of primes $p$.  By base change, the inert formula
$a_\fP = a_p^2 - 2p < 0$ holds whenever $|a_p| < \sqrt{2p}$, which
occurs for infinitely many $p$.
\end{remark}


% ===========================================================================
\section{Parsimony and Thurston Frequencies}\label{sec:parsimony}
% ===========================================================================

A single-step eversion achieves the sign flip but has only one corrugation
frequency.  For a visibly structured eversion with Thurston multi-frequency
corrugation, we use a short multi-prime sequence.

\begin{theorem}[Parsimonious eversion]\label{thm:parsimony}
The sequence $(\fP_5, \fP_3, \fP_{31})$ with $p = 5$ (ramified,
$a_\fP = 1$, $\Nm = 5$), $p = 3$ (inert, $a_\fP = -5$, $\Nm = 9$),
$p = 31$ (split, $a_\fP = 7$, $\Nm = 31$) achieves:
\begin{enumerate}[label=(\roman*)]
\item Sign flip: $\prod a_{\fP_k} = 1 \cdot (-5) \cdot 7 = -35 < 0$.
\item Three distinct corrugation frequencies: $k_\fP \in \{5, 9, 31\}$.
\item One prime of each splitting type (ramified, inert, split).
\end{enumerate}
\end{theorem}

The Thurston corrugation parameters at each step are:
\begin{center}
\begin{tabular}{llcccl}
\toprule
Step & Prime & $\Nm(\fP)$ & $a_\fP$ & $A_\fP = |a_\fP|/\Nm(\fP)$ & Type \\
\midrule
1 & $\fP_5$  & 5  & $+1$  & 0.200 & ramified \\
2 & $\fP_3$  & 9  & $-5$  & 0.556 & inert (sign flip) \\
3 & $\fP_{31}$ & 31 & $+7$ & 0.226 & split \\
\bottomrule
\end{tabular}
\end{center}

In the theory-true pipeline, the corrugation is not applied as an
explicit vertex displacement.  Instead, the Hecke operator $T_{\fP_k}$
is expressed as a $\Cl(3,1)$ multivector via $\rho^{-1}$, applied as a
full 16-component left-multiplication on the volumetric field, and then
propagated spatially by the Dirac leapfrog.  The resulting oscillations
in the witness scalar field at frequency $\Nm(\fP_k)$ correspond to the
Thurston corrugation formula
\begin{equation}\label{eq:thurston}
  F_k(x) \sim F_{k-1}(x) + \frac{A_{\fP_k}}{k_{\fP_k}}
    \sin\bigl(k_{\fP_k}\,\theta(x)\bigr)\,\mathbf{u}(x),
\end{equation}
but emerge from the arithmetic rather than being imposed post-hoc.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    >=Latex,
    every node/.style={font=\small},
    primestep/.style={circle,draw=accentBlue!75,fill=accentBlue!10,
                 thick,inner sep=2pt,minimum size=34pt},
    flip/.style={circle,draw=red!70,fill=red!10,
                 thick,inner sep=2pt,minimum size=34pt},
    arr/.style={->,thick,accentGray!85,line width=0.9pt},
    lab/.style={font=\footnotesize,text=accentGray},
    valpos/.style={font=\footnotesize,text=accentBlue!75},
    valneg/.style={font=\footnotesize,text=red!75}
  ]
  \draw[accentGray,thick,->] (-0.6,0) -- (10.8,0) node[right,lab] {step $k$};
  \draw[accentGray,thick,->] (-0.6,0) -- (-0.6,2.6) node[above,lab] {$\prod_{j\le k} a_{\fP_j}$ (cuspidal product)};
  \draw[accentGray,dashed] (-0.6,1.3) -- (10.4,1.3) node[right,lab] {$0$ (sign-flip threshold)};
  \coordinate (S0) at (0.0,2.2);
  \coordinate (S1) at (2.8,2.2);
  \coordinate (S2) at (5.6,0.3);
  \coordinate (S3) at (8.4,0.3);
  \node[primestep] at (S0) {$+1$};
  \node[primestep] at (S1) {$+1$};
  \node[flip] at (S2) {$-5$};
  \node[primestep,draw=red!70,fill=red!10] at (S3) {$-35$};
  %% Step arrows with prime annotations
  \draw[arr] (S0) -- node[valpos,above] {$\fP_5\;(\Nm=5,\,a=+1)$} (S1);
  \draw[arr] (S1) -- node[valneg,sloped,above] {$\fP_3\;(\Nm=9,\,a=-5)$} (S2);
  \draw[arr] (S2) -- node[valneg,above] {$\fP_{31}\;(\Nm=31,\,a=+7)$} (S3);
  %% Below: corrugation icons (snake decorations standing in for sinusoidal
  %% corrugations of frequency k and amplitude A; the visual encoding is
  %% schematic but ranked correctly: more zigs == higher k, taller zigs == higher A)
  \begin{scope}[yshift=-1.8cm]
    %% k=5, A=0.20 (mild amplitude, low frequency)
    \draw[accentBlue,thick,decorate,
          decoration={snake,amplitude=2pt,segment length=6pt}]
          (0,0) -- (2,0);
    \node[lab] at (1.0,-0.5) {corrugation $k=5$, $A=0.20$};
    %% k=9, A=0.56 (large amplitude, higher frequency, sign-flip step)
    \draw[red,thick,decorate,
          decoration={snake,amplitude=4pt,segment length=4pt}]
          (3,0) -- (5,0);
    \node[lab,text=red!75] at (4.0,-0.5) {$k=9$, $A=0.56$ (sign-flip step)};
    %% k=31, A=0.23 (mild amplitude, very high frequency)
    \draw[accentBlue,thick,decorate,
          decoration={snake,amplitude=2pt,segment length=2.5pt}]
          (6,0) -- (8,0);
    \node[lab] at (7.0,-0.5) {$k=31$, $A=0.23$};
  \end{scope}
\end{tikzpicture}
\caption{The three-prime eversion sequence
$(\fP_5,\fP_3,\fP_{31})$ in cuspidal-product coordinates.  Top
panel: the running cuspidal product
$\prod_{j\le k} a_{\fP_j}$ starts at $+1$, stays positive after the
ramified prime $\fP_5$ ($a=+1$), \emph{flips} sign at the inert
prime $\fP_3$ ($a=-5$), and lands at $-35$ after the split prime
$\fP_{31}$ ($a=+7$).  The sign change of the product is the
arithmetic signature of sphere eversion: the cuspidal component of
the coherence state has reversed orientation.  Bottom panel: the
Thurston corrugation icons at each step, with frequencies
$k = \Nm(\fP) \in \{5, 9, 31\}$ and amplitudes
$A_\fP = |a_\fP|/\Nm(\fP) \in \{0.20,\,0.56,\,0.23\}$.  The
sign-flip step is drawn inverted (red): it is the moment Thurston's
construction ``turns the sphere inside out.''  All three frequencies
co-exist in the final volumetric field via the Hecke composition
$T_{\fP_{31}} T_{\fP_3} T_{\fP_5}\,\psi_0$.}
\label{fig:eversion-trajectory}
\end{figure}

% ===========================================================================
\section{The Thurston Frequency Identity}\label{sec:thurston-identity}
% ===========================================================================

\begin{proposition}[Natural frequency limit]
As $p \to \infty$ through primes of $\QQ$, the corrugation frequencies
$k_\fP = \Nm(\fP)$ satisfy $k_\fP \to \infty$.  The Thurston formal limit
$k \to \infty$ (in which the corrugation becomes a $C^1$-dense family of
immersions) is realized as the limit over $\Spec(\cO_K)$ ordered by norm.
\end{proposition}

\begin{proof}
For split primes, $\Nm(\fP) = p \to \infty$.  For inert primes,
$\Nm(\fP) = p^2 \to \infty$.  Hence $\min(k_\fP) \to \infty$ as the
prime-ideal sequence extends.
\end{proof}

This identification is the key structural claim: the Thurston
corrugation method for sphere eversion is not an external construction
applied to the arithmetic---it IS the arithmetic, with the prime-ideal
norms providing the natural frequency scale.


% ===========================================================================
\section{Computational Verification}\label{sec:verification}
% ===========================================================================

All claims are verified by 29 tests (\texttt{test\_eversion.py}):

\begin{center}
\begin{tabular}{lcp{7cm}}
\toprule
Test class & Count & Verifies \\
\midrule
\texttt{TestPathDynamics}     & 3 & Hecke commutativity, exact product formula, path length \\
\texttt{TestSignFlip}         & 3 & Single negative $a_\fP$ flips; positive does not; mixed sequence matches product sign \\
\texttt{TestParsimony}        & 4 & Achieves eversion, bounded length, includes negative $a_\fP$, sorted by frequency \\
\texttt{TestRhoBridge}        & 5 & $\rho$ round-trip (random \& Hecke), $\rho^{-1}(T_\fP)$ preserves eigenvalues, left-mult $=$ geometric product, atlas seed $\propto\rho^{-1}(\psi\psi^T)$ \\
\texttt{TestAcousticPropagation} & 3 & Dirac leapfrog energy bounded; Hecke populates all 16 blades; Hecke + Dirac produces spatially varying field \\
\texttt{TestWitnessSkinTopology}  & 4 & Icosphere $\chi = 2$; marching cubes produces mesh; no degenerate triangles; intermediate topology differs \\
\texttt{TestMarchingCubes}    & 3 & Spherical field, empty field, vertex deduplication \\
\texttt{TestSmaleInvariant}   & 4 & Eversion: invariant 1; no flip: 0; double flip: 0; triple flip: 1 \\
\bottomrule
\end{tabular}
\end{center}

All 29 tests pass.  The witness surface is extracted via marching cubes
on the $64^3$ (or $32^3$) witness scalar grid, producing variable-topology
meshes that can self-intersect---capturing the Morin-type intermediate
that the field passes through during the cuspidal sign flip.


% ===========================================================================
\section{Interpretation}\label{sec:interpretation}
% ===========================================================================

\subsection{The eversion as arithmetic propagation}

The traditional approach to sphere eversion constructs a global regular
homotopy $F_t: S^2 \to \RR^3$ via either the Morin halfway model or
Thurston's corrugation method.  Both require continuous deformation of
an embedded surface through self-intersections.

The information-acoustic approach replaces continuous surface deformation
with discrete Hecke propagation on a 4-dimensional homological state space.
The ``surface'' is a derived object---the level set of a witness scalar
field in a volumetric Cl(3,1) substrate---not a primary geometric entity.

\subsection{Why the signature matters}

The Lorentzian signature $(3,1)$ of $\Cl(3,1)$ is not a choice: it is
the Eisenstein--cuspidal splitting of $H_1$ (Paper~VI).  The three
spacelike directions are the Eisenstein boundary modes; the single
timelike direction is the cuspidal interior mode.  The eversion is
propagation along the timelike axis: the cuspidal component flips sign,
which reverses the inside/outside orientation of the witness sphere.

\subsection{The parsimonious eversion principle}

Informational parsimony demands that the eversion use the minimum
arithmetic structure: one prime per splitting type, ordered by norm,
truncated at the first sign flip.  The resulting 3-prime sequence
$(5, 3, 31)$ is not arbitrary---it is the unique shortest sequence
drawn from the first primes of $\cO_K$ that achieves multi-frequency
corrugation with an odd number of cuspidal sign flips.

\subsection{Forward reference: the eversion law on byte chains}

The cuspidal sign-flip mechanism that drives geometric eversion
generalizes to a per-chain parity law on arbitrary blade sequences
in $\Cl(3,1)$.  In Paper~VIII (Layer~3) this is sharpened to the
\emph{eversion law}
\[
  M_{T_{13}}(c) = \varepsilon(c) \cdot M_{T_{31}}(c),
  \quad
  \varepsilon(c) = (-1)^{\lfloor c_2/2 \rfloor + \lfloor c_3/2 \rfloor},
\]
where $c_i$ counts occurrences of generator $e_i$ across a blade
chain $c$.  The same $\varepsilon(c)$ then forms one of the three
identities of the Combined Reading Lemma
(Paper~VIII, Section~12), which proves that engine state and NSM
state decompose into orthogonal halves of $\Cl(3,1)$ for every
byte chain; the witness pair $(\mathsf{be}, \mathsf{know}) =
\mathrm{span}\{e_0, e_{15}\}$ is the unique chirality-invariant mediator
between them.  Sphere eversion in this paper is thus the geometric
realization of the same single-bit parity operation that drives
voice swap in language.


% ===========================================================================
\section*{Acknowledgments}
% ===========================================================================

\begin{sloppypar}
Computational verification uses the \texttt{substrate-sim/algebra}
codebase.  Core modules: \texttt{eversion} (path dynamics, parsimony,
Smale invariant), \texttt{acoustic\_lift} ($64^3$ Cl(3,1) atlas, Dirac
leapfrog, witness scalar), \texttt{hilbert\_homology} (4-dim $H_1$ and
spectral Hecke), \texttt{quaternion\_bridge} ($\Cl(3,1) \to \End(H_1)$
representation, $\rho^{-1}$, left-multiplication),
\texttt{e11\_point\_count} (first-principles eigenvalues).
29 tests, 0 failures.
\end{sloppypar}


\begin{thebibliography}{99}

\bibitem{paper6}
Paper~VI: The Operator-Level Bridge.  From Peirce Decomposition to Prime
Splitting via the Quaternion--Clifford Identification.

\bibitem{thurston1974}
W.~Thurston, \emph{Corrugations}, lecture notes, Princeton, 1974.

\bibitem{smale1958}
S.~Smale, \emph{A classification of immersions of the two-sphere},
Trans.\ Amer.\ Math.\ Soc.\ \textbf{90} (1958), 281--290.

\bibitem{morin1978}
B.~Morin, \emph{Formes canoniques des singularit\'es d'une application
diff\'erentiable}, C.~R.\ Acad.\ Sci.\ Paris \textbf{260} (1965), 5662--5665.

\bibitem{max1977}
N.~Max, \emph{Turning a Sphere Inside Out} (film), International Film
Bureau, 1977.

\end{thebibliography}

\end{document}
